---
title: "EcoPlotsAPI"
author: "Arun Singh Ramesh^[TERN, a.singhramesh@uq.edu.au]"
date: "28/09/2023"
output:
  html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 8
    fig_width: 12
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
    self_contained: true
    github_document: true
output_dir: "docs"
documentclass: article
fontsize: 24pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## EcoPlots API example

The example code shown here is designed to give you an experience at downloading, analyzing and visualising a dataset that is archived on the TERN EcoPlots https://ecoplots-test.tern.org.au/. 
EcoPlots contains plot-based ecology data from different sources to enable integrated search and access to data based on different jurisdiction, data sources, feature types, parameters and temporal extent. 

# General preparations
### **Set your working directory**

This gets you started with storing your files and r scripts.

```{r working directory}
#setwd("C:/Users/uqasin21") # Note that your path will be different
```
### **Load libraries**
If you do not have the following packages downloaded, you can install them using the function `install.packages`, with the name of the package as an argument in quotations.


```{r load, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(httr)
library(ggpubr)
library(gridExtra)
```

# **Load and explore the data**

At this point, you should have your Ecoplots search query from the EcoPlots API Query dash board (https://ecoplots-test.tern.org.au/discovery).

***Important*** Make sure to download your API key from the website (https://account-test.tern.org.au/authenticated_user/apikeys)
Once you have your API code extracted you can start using the query request

```{r API, results='hide', eval=TRUE, message=FALSE, warning=FALSE}
headers = c(
  'X-Api-Key' = 'your-api-key',
  'Content-Type' = 'application/json'
)

body = '{
  "query": {
    "dataset": [
      "http://linked.data.gov.au/dataset/bdbsa"
    ],
    "observed_property": [
      "http://linked.data.gov.au/def/tern-cv/fc2dcb5b-94a5-440f-9104-efaa640246cc",
      "http://linked.data.gov.au/def/tern-cv/56195246-ec5d-4050-a1c6-af786fbec715",
      "http://linked.data.gov.au/def/tern-cv/800cf33b-1c44-435d-89bd-5d80e3e8aba1"
    ],
    "feature_type": [
      "http://linked.data.gov.au/def/tern-cv/226a127b-7248-476a-84ad-0412a5e3af52"
    ]
  }
}';

res <- VERB("POST", url = "https://ecoplots-test.tern.org.au/api/v1.0/data/bdbsa?dformat=csv", body = body, add_headers(headers))

cat(content(res, 'text'))

df <- read.table(text = content(res, 'text'), sep =",", header = TRUE, stringsAsFactors = FALSE)


```

# **Explore your data frame and do simple stats**

First step is to explore the data frame and organise it to suite your 
```{r  stats0, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
#data exploration and wrangling
head(df)
tail(df)
summary(df) 

#convert your categorical values into factor levels
df$bodyMass_gram<- as.numeric(df$bodyMass_gram)
df$scientificName<- as.factor(df$scientificName)

# visualise the general distribution of your response variable (body mass)
hist(df$bodyMass_gram, breaks = 150)

```

It looks like the data has many vectors either as a categorical or numeric vector. We need to organise the dataframe and store the categorical values as a factor. 

```{r  stats1, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}

#convert your numerical values to numeric and categorical values into factor levels
df$bodyMass_gram<- as.numeric(df$bodyMass_gram)
df$scientificName<- as.factor(df$scientificName)

# visualise the general distribution of your response variable (body mass)
hist(df$bodyMass_gram, breaks = 150)

```
It looks like majority of the dataset contains body mass distribution of mammals below 300g. 
Here we subset the data frame to represent body mass below 50g for example and then see if the distribution is normal.


```{r  stats2, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}
#subset
dflow<- subset(df, bodyMass_gram<=50)
hist(dflow$bodyMass_gram, breaks = 100)

```

Now we have a new dataframe with 6703 hits. And the dataset appears a better representation of body size

# **Data Visualisation**
For the first instance, let us visualise a relation between body mass with elevation

```{r  vis1, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}

plot(dflow$bodyMass_gram~dflow$altitude_Metre)

```

As you can see the data is scattered all over with higher point clouds at lower elevation sites. 
Let visualise with a regression line and see how the slope of the relationship changes with elevation.

```{r  vis2, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}

#fancy data visualisation
library(ggpubr)

l1<-ggscatter(dflow, x = "altitude_Metre", y = "bodyMass_gram", 
              color = "#C33453", pch=19, size=5,alpha=0.8)+ggtitle("")+ 
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")),label.y = 45, label.x = 700, size=4)+stat_smooth(se=TRUE, formula = y~x, color="grey")+
  stat_regline_equation(label.y =47 , label.x = 700, size=4)+
  xlab(expression(Elevation~(meter~a.s.l.)))+ ylab(expression(Body~Mass~("g")))+ theme_bw()

l1+ theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), 
          axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
          legend.text = element_text(size = 12, face ="bold", colour ="black"), 
          legend.position = "top", axis.title.y = element_text(face = "bold", size = 14), 
          axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
          legend.title = element_text(size = 14, colour = "black", face = "bold"), 
          panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
          legend.key=element_blank())
```

As you can see there is a weak but negative slope, suggesting that most large mammals are found in lower elevation sites.

# **Body mass summary at Site/Species level**

Now that we have a relation of body mass with elevation. Let us normalise the data and calculate the trends across a particular bioregion. In this example we use two contrasting biomes- a coast versus an inland desert.

```{r  vis3, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}

#aggregate biomass by Species, Project and Site
dflow$projectTitle<- as.factor(dflow$projectTitle)
dflow$siteName<- as.factor(dflow$siteName)

library(tidyverse)
library(gridExtra)

dflow%>%
  group_by(siteName,projectTitle,scientificName)%>%
  summarize(bodyMass_gram=sum(bodyMass_gram,na.rm=TRUE))->summed_species_level_biomass

# let us see the body mass averaged for each species 
with(summed_species_level_biomass, tapply(bodyMass_gram, c(scientificName), mean))

#subset data for the project on Eyre Peninsula (coast) in SA

Eyre_Peninsula<- subset(summed_species_level_biomass, projectTitle=="Southern Eyre Peninsula")

#plot species biomass by Project location (one in coast versus inland desert)
#Lets look at the Coast
plot1<-ggplot(Eyre_Peninsula, aes(x = scientificName, y = bodyMass_gram))+ggtitle("Eyre Peninsula")+geom_col(position="dodge")+labs(x="Species",y="Body mass (g)")
p1<- plot1+theme_bw() +theme(axis.text.x = element_text(angle = 45, hjust = 1))
p1


```

There appears to be fewer species recorded from this Coast with the species <i> Mus musculus</i> accounting for the highest body mass.

Let us see what body sizes we observe for mammals within the Great Sandy desert

```{r  vis4, results='markdown', eval=TRUE, message=FALSE, warning=FALSE}

#subset data for the Project on Sandy Desert in SA

Sandy_desert<-subset(summed_species_level_biomass,projectTitle=="Sandy Desert")

#plot species biomass by Project location (inland desert)

#Sandy desert
plot2<-ggplot(Sandy_desert, aes(x = scientificName, y = bodyMass_gram))+geom_col(position="dodge")+labs(x="Species",y="Body mass (g)")
p2<- plot2+theme_bw() +theme(axis.text.x = element_text(angle = 45, hjust = 1))
p2

```

# Summary
* We found variation in body mass with elevation 
and 
* variation in body mass of mammals from coast to the inland deserts in the SA region.


# References

[TERN EcoPlots- Terrestrial Ecosystem Research Network.](https://ecoplots.tern.org.au/)

[The Biological Databases of South Australia (BDBSA)](https://www.environment.sa.gov.au/topics/science/information-and-data/biological-databases-of-south-australia)
